---
title: "Plotting hubverse formatted output"
description: >
   In this vignette you will learn how to use
   `forecasttools` to make quick visualizations
   of hubverse formatted output for spot-checking.
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Plotting hubverse formatted output}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  dev = "svg",
  fig.ext = "svg",
  fig.height = 5,
  fig.width = 5
)
```


After generating a hubverse formatted forecast, it is good to inspect the output to make sure that the results make sense. To help you, `forecasttools` provides convenience functions for making timeseries plots from hubverse-format output.

First, let's load `forecasttools`.
```{r, message = FALSE}
library(forecasttools)
```

# Quantile timeseries
Much hubverse formatted output is organized into quantiles. An easy spotcheck plot shows how these quantiles evolve over the forecast horizon. We can make it using the `plot_hubverse_quantiles()` function. There is only one mandatory argument: the path to a properly hubverse-formatted `.csv` file. Let's plot some inflenza forecasts submitted to the 2023-24 FluSight Challenge by the `cfarenewal-cfaepimlight` team for the 2024-04-06 reference date:

```{r, message = FALSE, echo = c(2, 4)}
# nolint start
path_to_formatted_forecast <- "https://raw.githubusercontent.com/cdcepi/FluSight-forecast-hub/refs/heads/main/model-output/cfarenewal-cfaepimlight/2024-04-06-cfarenewal-cfaepimlight.csv"
# nolint end
plots <- plot_hubverse_quantiles(path_to_formatted_forecast)
```

`plot_hubverse_quantiles()` returns a list of all the plots generated. By default, the list names (keys) are US Postal Service style two-letter abbreviations. Let's look at the national plot:

```{r}
plots[["US"]]
```

We can also look at the plot for Colorado:

```{r}
plots[["CO"]]
```

Of course, you may not wish to generate plots for all locations in your hubverse-formatted file at once. `plot_hubverse_quantiles()` takes an optional `locations` argument that allows you to plot only a subset. For example, let's plot the ["Four Corners" states](https://en.wikipedia.org/wiki/Four_Corners):

```{r, message = FALSE}
four_corners <- plot_hubverse_quantiles(path_to_formatted_forecast,
  locations = c("AZ", "CO", "NM", "UT")
)

## display New Mexico
four_corners[["NM"]]
```

Many hubs provide "target data" or "truth data" of observed values of the forecasting target, and so `plot_hubverse_quantiles()` optionally allows you to plot this alongside the forecast data. Since this target data often goes back years, it is useful to set a cutoff using the `start_date` argument. Here, we'll start in  December 2023.

```{r, message = FALSE, echo = c(2, 4:7)}
# nolint start
target_data_path <- "https://raw.githubusercontent.com/cdcepi/FluSight-forecast-hub/04e884dce942dd3b8766aee3d8ff1c333b4fb6fa/target-data/target-hospital-admissions.csv"
# nolint end
plot_hubverse_quantiles(path_to_formatted_forecast,
  locations = "US",
  observed_data_path = target_data_path,
  start_date = "2023-12-01"
)
```

The function provides some basic customization of plotted lines and points via the `linewidth`, `pointsize`, `forecast_linecolor`, `forecast_pointcolor`, `obs_pointcolor`, and `obs_linecolor` arguments. It also defaults to plotting on a log10-scale y-axis, but this can be changed by passing a different string to `y_transform`; any valid value for the `transform = ` argument of `ggplot2::scale_y_continuous()` can be passed.

```{r, message = FALSE}
## plot forecast data in green, with smaller points and
## lines, and plot on a linear scale
my_custom_plot <- plot_hubverse_quantiles(path_to_formatted_forecast,
  locations = "US",
  observed_data_path = target_data_path,
  start_date = "2023-12-01",
  forecast_linecolor = "darkgreen",
  forecast_pointcolor = "darkgreen",
  pointsize = 1,
  linewidth = 1,
  ytrans = "identity"
)
my_custom_plot[["US"]]
```

For further customization, you can modify the resulting ggplot objects, as you would a regular ggplot. For example, we can convert the above plot to the classic ggplot2 theme.

```{r, message = FALSE}
library(ggplot2)
my_custom_plot[["US"]] + theme_classic()
```
