---
title: "Retrieving NHSN hospital respiratory data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Retrieving NHSN hospital respiratory data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
httptest::start_vignette("pull-nhsn")
```

In this vignette you will learn how use `forecasttools` to pull data from the [National Healthcare Safety Network (NHSN)](https://www.cdc.gov/nhsn/index.html) [Weekly Hospital Respiratory Data (HRD)](https://data.cdc.gov/Public-Health-Surveillance/Weekly-Hospital-Respiratory-Data-HRD-Metrics-by-Ju/ua7e-t2fy/about_data) dataset. This dataset contains counts of epiweekly influenza and COVID-19 hospital admissions by U.S. state, among other quantities.

We will use the [Socrata Open Data API (SODA)](https://dev.socrata.com/) API endpoint on [`data.cdc.gov`](https://data.cdc.gov). `data.cdc.gov` also provides a [browser view of the data](https://data.cdc.gov/Public-Health-Surveillance/Weekly-Hospital-Respiratory-Data-HRD-Metrics-by-Ju/ua7e-t2fy) with links to download data (e.g. as `.csv` or `.tsv` files), but when building pipelines, we wish to operate programmatically.


## API Key
First, you'll want to go to `data.cdc.gov` and request an API token. This is not strictly required, but it is generally considered polite, and it will speed up your data requests (the polite get served first!).

To request a token, navigate to [`data.cdc.gov`'s developer settings page](https://data.cdc.gov/profile/edit/developer_settings). You will be prompted to log in. If you have CDC credentials, you can use those. Otherwise, you can [sign up for an account](https://data.cdc.gov/signup) with Tyler Data and Insights (the contractor that manages `data.cdc.gov`). Once logged in, navigate to the [developer settings page](https://data.cdc.gov/profile/edit/developer_settings), click the "Create new API key" button and follow the prompts.

Make sure to record your secret key somewhere safe but _not_ tracked by Git (e.g. a `.gitignore`-ed `secrets.toml` file.

One place to store secrets is as [environment variables](https://en.wikipedia.org/wiki/Environment_variable). If you don't provide one explicitly, `pull_nhsn()` looks for a valid NHSN API key ID in an environment variable named `NHSN_API_KEY_ID` and a corresponding secret in an environment variable named `NHSN_API_KEY_SECRET`.

## Getting all the data
Our workhorse function for getting NHSN data is called simply `pull_nhsn()`. If you provide it with no arguments, it will simply fetch you the entire dataset as a `tibble::tibble()`, up to the specified maximum number of rows (default 10,0000).

Yyou can increase by setting the `limit` argument of `pull_nhsn()` a larger value. To protect you against accidentally pulling incomplete datasets, `pull_nhsn()` errors by default if the number of rows retrieved hits the `limit`. You can suppress that error by setting `error_on_limit = FALSE`.

We'll provide the `api_key_id` and `api_key_secret` arguments explicitly. If you omit those (and valid values can't be found in the default environment variables), `forecasttools` will warn you that the data request will be lower priority and less polite.

```{r}
my_api_key_id <- "<YOUR API KEY ID HERE>"
my_api_key_secret <- "<YOUR API KEY SECRET HERE>"
```

```{r, include = FALSE}
# silently reconcile env var secrets and mock var secrets
# so that the vignette works as expected with both mock
# and real requests provided that real requests are made with
# real secrets (and mock requests do not require real secrets).
env_id <- Sys.getenv("NHSN_API_KEY_ID")
env_secret <- Sys.getenv("NHSN_API_KEY_SECRET")

if (env_id == "") {
  Sys.setenv("NHSN_API_KEY_ID") <- my_api_key_id
} else {
  my_api_key_id <- env_id
}

if (env_secret == "") {
  Sys.setenv("NHSN_API_KEY_SECRET") <- my_api_key_secret
} else {
  my_api_key_secret <- env_secret
}
```

```{r}
library(forecasttools)
library(dplyr)
all_data <- pull_nhsn(
  limit = 1000,
  api_key_id = my_api_key_id,
  api_key_secret = my_api_key_secret,
  error_on_limit = FALSE
)

all_data
```

## Getting only relevant rows and columns
This is a huge table, and it can take a fair amount of time to download. We can speed things up by only requesting a subset of the data via a "query". These queries are a way to ask `data.cdc.gov` only for the rows and columns we care about, which will speed up our download.

To automate repetitive tasks, there are pre-defined ways to apply some common queries.

### Date limits
`pull_nhsn()` takes `start_date` and `end_date` arguments. If these are defined, it will query the dataset for only rows that fall between those dates, inclusive

### Column limits
`pull_nhsn()` takes a `columns` argument, which defaults to `NULL`. If `columns` is `NULL`, `pull_nhsn()` pulls all columns. Otherwise, it pulls the columns `jurisdiction`, `weekendingdate`, and any columns enumerated in `columns`.

### An example
As an example, let's pull only epiweekly incident influenza hospitalizations between 1 Jan 2023 and 31 December 2023, inclusive. Here, we'll let `pull_nhsn()` look for an API key and secret in our environment variables.

```{r}
influenza_2023 <- pull_nhsn(
  columns = c("totalconfflunewadm"),
  start_date = "2023-01-01",
  end_date = "2023-12-31"
)

influenza_2023
```

By default, `pull_nhsn()` returns its output sorted first by `jurisdiction`, then by `weekendingdate`. You can configure this differently via the `order_by` keyword argument.

```{r}
influenza_by_date <- pull_nhsn(
  columns = c("totalconfflunewadm"),
  start_date = "2023-01-01",
  end_date = "2023-12-31",
  order_by = c("weekendingdate", "jurisdiction")
)

influenza_by_date
```

```{r, include=FALSE}
httptest::end_vignette()
```
